# Data Analysis

Opal Patronage

<https://opendata.transport.nsw.gov.au/data/dataset/opal-patronage>

January 2020

The structure of the data set

```{r}
library(ggplot2)
library(dplyr)

data <- read.table("data/opal_patronage/Opal_Patronage_20200101.txt",
                   header = TRUE,
                   sep = "|")

str(data)
```

There are 4 modes of transportation offered by Transport by NSW

```{r}
unique(data$mode_name)
```

The locations covered by each mode of transportion:

Train

```{r}
unique(filter(data, mode_name == "Train")$ti_region)
```

Bus

```{r}
unique(filter(data, mode_name == "Bus")$ti_region)
```

Ferry

```{r}
unique(filter(data, mode_name == "Ferry")$ti_region)
```

Light rail

```{r}
unique(filter(data, mode_name == "Light rail")$ti_region)
```

Combining all of the .txt files into one big .txt file

(DONT RUN) \~ FILE IS ALREADY CREATED

```{r}
library(dplyr)

folder_path <- "data/opal_patronage"
file_list <- list.files(path = folder_path, pattern = "\\.txt$", full.names = TRUE)

# Define column names
col_names <- c("trip_origin_date", "mode_name", "ti_region", "tap_hour", "Tap_Ons", "Tap_Offs")

# Function to read and clean a single file
read_clean_file <- function(file, skip_header = FALSE) {
  lines <- readLines(file, warn = FALSE)
  
  if (length(lines) < 2) return(NULL)  # Skip empty or header-only files
  
  if (skip_header) {
    lines <- lines[-1]
  }
  
  good_lines <- lines[sapply(lines, function(line) length(strsplit(line, "\\|")[[1]]) == 6)]
  
  if (length(good_lines) == 0) return(NULL)
  
  df <- read.table(text = good_lines, sep = "|", header = FALSE, stringsAsFactors = FALSE, colClasses = "character")
  colnames(df) <- col_names
  return(df)
}

# Read the first file
combined_data <- read_clean_file(file_list[1], skip_header = FALSE)

# Read and append the rest
for (file in file_list[-1]) {
  df <- read_clean_file(file, skip_header = TRUE)
  if (!is.null(df)) {
    combined_data <- bind_rows(combined_data, df)
  }
}

# Write the final result
write.table(combined_data,
            file = "data/opal_patronage/Opal_Patronage_Combined.txt",
            sep = "|",
            row.names = FALSE,
            col.names = TRUE,
            quote = FALSE)
```

Reading the combined .txt file

Finding:\
In some of the dataset, like 20220208, there is an UNKNOWN mode of transportation

```{r}
unique(combined_data$mode_name)
sum(combined_data$mode_name != "UNKNOWN")
nrow(combined_data)
```

Removing certain districts that are far from Sydney, NSW

```{r}
filtered_data <- combined_data[combined_data$ti_region %in% c("Parramatta", "Chatswood", "Macquarie Park", "North Sydney", "Strathfield", "Sydney CBD"), ]

# To confirm
unique(filtered_data$ti_region)
```

Combining all tap hours from 0 - 23 into per day

```{r}
# Replace "<50" with NA, then convert Tap_Ons and Tap_Offs to numeric
filtered_data$Tap_Ons <- as.numeric(gsub("<50", NA, filtered_data$Tap_Ons))
filtered_data$Tap_Offs <- as.numeric(gsub("<50", NA, filtered_data$Tap_Offs))

# Remove rows with NA values in Tap_Ons or Tap_Offs
filtered_data <- filtered_data[!is.na(filtered_data$Tap_Ons) & !is.na(filtered_data$Tap_Offs), ]

# Aggregate sums by trip_origin_date, mode_name, and ti_region
daily_totals_by_region <- aggregate(
  cbind(Tap_Ons, Tap_Offs) ~ trip_origin_date + mode_name + ti_region,
  data = filtered_data,
  FUN = sum
)

# Rename columns for clarity
colnames(daily_totals_by_region) <- c(
  "trip_origin_date",
  "mode_name",
  "ti_region",
  "Total_Tap_Ons",
  "Total_Tap_Offs"
)

str(daily_totals_by_region)
```

Confirming

```{r}
unique(filter(daily_totals_by_region, mode_name == "Train")$ti_region)
```

# Rainfall Data set

Reading rainfall data (choosing Sydney Botanic Garden meteorology station)

```{r}
rain <- read.csv("data/rainfall/sydney_botanic_garden.csv",
                 header = TRUE,)

str(rain)
```

Filtering the date so it starts on 1st January 2020

```{r}
filter(rain, Year >= 2020)

```

Merging the Opal Patronage and rainfall datasets into one

```{r}
library(dplyr)
library(readr)
library(lubridate)

# 1. Load and combine rainfall data (same as before)
rainfall_folder <- "data/rainfall"
rainfall_files <- list.files(rainfall_folder, pattern = "\\.csv$", full.names = TRUE)

rainfall_list <- lapply(rainfall_files, function(file) {
  region <- tools::file_path_sans_ext(basename(file))
  
  df <- read_csv(file, show_col_types = FALSE)
  
  df <- df %>%
    mutate(
      # keep ti_region exactly as filename (with underscores)
      ti_region = tolower(trimws(region)),
      date = as.Date(paste(Year, Month, Day, sep = "-"))
    ) %>%
    select(ti_region, date, Rainfall_amount = `Rainfall amount (millimetres)`)
  
  return(df)
})

rainfall_data <- bind_rows(rainfall_list)

# 2. Prepare Opal data, but replace spaces with underscores to match rainfall_data
daily_totals_by_region <- daily_totals_by_region %>%
  mutate(
    # lowercase + trim + replace spaces with underscores
    ti_region = tolower(trimws(ti_region)),
    ti_region = gsub(" ", "_", ti_region),
    trip_origin_date = as.Date(trip_origin_date)
  )

# 3. Join on ti_region and date
big_data <- daily_totals_by_region %>%
  left_join(rainfall_data, by = c("ti_region", "trip_origin_date" = "date"))

# Check the result
head(big_data)
summary(big_data$Rainfall_amount)




```

```{r}
unique(filter(big_data, mode_name == "Train")$ti_region)
```

Plotting tap ins vs rainfall

```{r}
library(dplyr)
library(ggplot2)

big_data %>%
  filter(tolower(mode_name) == "train", tolower(ti_region) == "sydney_cbd") %>%
  ggplot(aes(x = Rainfall_amount, y = Total_Tap_Ons)) +
  geom_point(alpha = 0.6, color = "blue") +
  labs(
    title = "Train Tap-Ons vs Rainfall in Sydney CBD",
    x = "Rainfall (mm)",
    y = "Total Opal Tap-Ons"
  ) +
  theme_minimal()



```

### Train (Tap ins vs Rainfall)

```{r}
library(dplyr)
library(ggplot2)

big_data %>%
  filter(tolower(mode_name) == "train") %>%
  ggplot(aes(x = Rainfall_amount, y = Total_Tap_Ons)) +
  geom_point(alpha = 0.6, color = "blue") +
  facet_wrap(~ ti_region, scales = "free") +  # One plot per region, scales free for each
  labs(
    title = "Train Tap-Ons vs Rainfall by Region",
    x = "Rainfall (mm)",
    y = "Total Opal Tap-Ons"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 8)  # smaller facet labels if many regions
  )

```

Checking correlation

```{r}
library(dplyr)

# Filter train data (example: Sydney CBD)
train_sydney <- big_data %>%
  filter(tolower(mode_name) == "train", ti_region == "sydney_cbd") %>%
  filter(!is.na(Rainfall_amount) & !is.na(Total_Tap_Ons))

# Calculate Pearson correlation
correlation <- cor(train_sydney$Rainfall_amount, train_sydney$Total_Tap_Ons)

print(correlation)

```

Bus

```{r}
big_data %>%
  filter(tolower(mode_name) == "bus") %>%
  ggplot(aes(x = Rainfall_amount, y = Total_Tap_Ons)) +
  geom_point(alpha = 0.6, color = "red") +
  facet_wrap(~ ti_region, scales = "free") +  # One plot per region, scales free for each
  labs(
    title = "Bus Tap-Ons vs Rainfall by Region",
    x = "Rainfall (mm)",
    y = "Total Opal Tap-Ons"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 8)  # smaller facet labels if many regions
  )
```

Ferrry

```{r}
big_data %>%
  filter(tolower(mode_name) == "ferry") %>%
  ggplot(aes(x = Rainfall_amount, y = Total_Tap_Ons)) +
  geom_point(alpha = 0.6, color = "green") +
  facet_wrap(~ ti_region, scales = "free") +  # One plot per region, scales free for each
  labs(
    title = "Ferry Tap-Ons vs Rainfall by Region",
    x = "Rainfall (mm)",
    y = "Total Opal Tap-Ons"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 8)  # smaller facet labels if many regions
  )
```

Light rail

```{r}
big_data %>%
  filter(tolower(mode_name) == "light rail") %>%
  ggplot(aes(x = Rainfall_amount, y = Total_Tap_Ons)) +
  geom_point(alpha = 0.6, color = "purple") +
  facet_wrap(~ ti_region, scales = "free") +  # One plot per region, scales free for each
  labs(
    title = "Ferry Tap-Ons vs Rainfall by Region",
    x = "Rainfall (mm)",
    y = "Total Opal Tap-Ons"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 8)  # smaller facet labels if many regions
  )
```

Greater Sydney Map

<https://www.abs.gov.au/statistics/standards/australian-statistical-geography-standard-asgs-edition-3/jul2021-jun2026/access-and-downloads/digital-boundary-files>

```{r}
library(sf)
library(dplyr)
library(ggplot2)

# Load shapefile (make sure all shapefile files are in the folder)
sa2 <- st_read("data/sydney_map/")

# Filter Greater Sydney districts
sydney_sa2 <- sa2 %>% filter(GCC_NAME21 == "Greater Sydney")

# Plot the Greater Sydney SA2 map
ggplot(sydney_sa2) +
  geom_sf(fill = "lightblue", color = "black", size = 0.3) +
  labs(title = "Greater Sydney Districts (SA2 Level)") +
  theme_minimal()


```

heme_minimal()

```{r}
library(sf)
library(ggplot2)

sa2 <- st_read("data/sydney_map/")

# Define bounding box (xmin, ymin, xmax, ymax) - coordinates roughly around Sydney metro area
bbox <- st_bbox(c(xmin = 150.9, ymin = -33.9, xmax = 151.3, ymax = -33.8), crs = st_crs(sa2))

# Crop polygons to bounding box
focus_area <- st_crop(sa2, bbox)

ggplot(focus_area) +
  geom_sf(fill = "lightblue", color = "black", size = 0.3) +
  labs(title = "Zoomed Map Near Central Sydney") +
  theme_minimal()

```

```{r}
library(sf)
library(dplyr)
library(ggplot2)

# Load SA2 shapefile
sa2 <- st_read("data/sydney_map/")

# Define the exact SA2 districts you want to zoom into
target_sa2 <- c("Sydney - Central", "North Sydney", "Parramatta", "Strathfield")

# Filter only those districts
focus_sa2 <- sa2 %>% filter(SA2_NAME21 %in% target_sa2)

# Join your Opal tap-ons data (make sure daily_totals_by_region exists)
sa2_opal_focus <- focus_sa2 %>%
  left_join(daily_totals_by_region, by = c("SA2_NAME21" = "ti_region"))

# Plot zoomed-in map
ggplot(sa2_opal_focus) +
  geom_sf(aes(fill = Total_Tap_Ons), color = "black", size = 0.2) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90") +
  labs(
    fill = "Total Tap-Ons",
    title = "Opal Tap-Ons Near Sydney Central and Neighbours",
    caption = "Data source: Opal Patronage & ABS"
  ) +
  theme_minimal()

unique(sa2$SA2_NAME21)


```

Using SA3 data because SA2 might be too detailed

```{r}
library(sf)
library(dplyr)
library(ggplot2)

# Load SA3 shapefile
sa3 <- st_read("data/sa3/SA3_2021_AUST_GDA2020.shp")

# Filter for Greater Sydney only
sydney_sa3 <- sa3 %>% filter(GCC_NAME21 == "Greater Sydney")

# Join Opal tap-ons data by matching district names
sa3_opal <- sydney_sa3 %>%
  left_join(daily_totals_by_region, by = c("SA3_NAME21" = "ti_region"))

# Plot choropleth map zoomed on Greater Sydney SA3 regions
ggplot(sa3_opal) +
  geom_sf(aes(fill = Total_Tap_Ons), color = "black", size = 0.2) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90") +
  labs(
    fill = "Total Tap-Ons",
    title = "Opal Train Tap-Ons in Greater Sydney (SA3 Districts)",
    caption = "Data source: Opal Patronage & ABS"
  ) +
  theme_minimal()

```

Zooming in on specific districts

```{r}
library(sf)
library(dplyr)
library(ggplot2)

# Load SA3 shapefile
sa3 <- st_read("data/sa3/SA3_2021_AUST_GDA2020.shp")

# Define bounding box near Sydney CBD (adjust these coordinates as needed)
bbox <- st_bbox(c(xmin = 150.95, ymin = -34.05, xmax = 151.15, ymax = -33.80), crs = st_crs(sa3))

# Crop SA3 polygons to this bbox
cropped_sa3 <- st_crop(sa3, bbox)

# Join Opal tap-ons data (assuming daily_totals_by_region has matching SA3_NAME21 in ti_region)
cropped_sa3_opal <- cropped_sa3 %>%
  left_join(daily_totals_by_region, by = c("SA3_NAME21" = "ti_region"))

# Plot cropped map
ggplot(cropped_sa3_opal) +
  geom_sf(aes(fill = Total_Tap_Ons), color = "black", size = 0.2) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90") +
  labs(
    fill = "Total Tap-Ons",
    title = "Opal Tap-Ons near Sydney CBD (custom bbox)",
    caption = "Data source: Opal Patronage & ABS"
  ) +
  theme_minimal()

```

Checking SA2 district names

```{r}
# Extract unique district names
district_names <- unique(sa2$SA2_NAME21)

# Optionally sort them
district_names <- sort(district_names)

# Write to a text file, one name per line
writeLines(district_names, "district_names_sa2.txt")


```

```{r}
library(sf)
library(dplyr)
library(stringr)
library(ggplot2)

# Your lookup table for Sydney SA2 -> Sydney Region
sydney_sa2_regions <- data.frame(
  SA2_NAME = c(
    "Darlinghurst", "Docklands", "Potts Point - Woolloomooloo", "Pyrmont",
    "Sydney (North) - Millers Point", "Sydney (South) - Haymarket",
    "Parramatta - North", "Parramatta - South", "Cremorne - Cammeray",
    "Crows Nest - Waverton", "Greenwich - Riverview", "Neutral Bay - Kirribilli",
    "North Sydney - Lavender Bay", "St Leonards - Naremburn", "Waverley - St Leonards",
    "Macquarie Park - Marsfield", "Strathfield - East", "Strathfield - West",
    "Strathfield South", "Artarmon", "Chatswood - East", "Chatswood (West) - Lane Cove North",
    "Lane Cove", "North Ryde - East Ryde", "Ryde - North", "Ryde - South",
    "West Ryde - Meadowbank"
  ),
  Sydney_Region = c(
    "Sydney CBD", "Sydney CBD", "Sydney CBD", "Sydney CBD", "Sydney CBD", "Sydney CBD",
    "Parramatta", "Parramatta", "North Sydney", "North Sydney", "North Sydney", "North Sydney",
    "North Sydney", "North Sydney", "North Sydney", "Macquarie Park",
    "Strathfield", "Strathfield", "Strathfield", "Chatswood", "Chatswood", "Chatswood",
    "Chatswood", "Chatswood", "Chatswood", "Chatswood", "Chatswood"
  ),
  stringsAsFactors = FALSE
)

# Helper function to clean names (lowercase, remove spaces and dashes)
clean_name <- function(x) {
  tolower(str_replace_all(x, "[- ()]", ""))
}

# Load SA2 shapefile (adjust path if needed)
sa2 <- st_read("data/sydney_map/SA2_2021_AUST_GDA2020.shp")

# Add cleaned name columns for matching
sa2 <- sa2 %>% mutate(clean_name = clean_name(SA2_NAME21))
sydney_sa2_regions <- sydney_sa2_regions %>% mutate(clean_name = clean_name(SA2_NAME))

# Filter SA2 polygons matching your Sydney SA2 names using cleaned names
sa2_sydney_regions <- sa2 %>%
  filter(clean_name %in% sydney_sa2_regions$clean_name) %>%
  left_join(sydney_sa2_regions, by = "clean_name")

# Dissolve polygons by Sydney_Region (union geometries)
sydney_region_polygons <- sa2_sydney_regions %>%
  group_by(Sydney_Region) %>%
  summarise(geometry = st_union(geometry)) %>%
  ungroup()

# Join Sydney_Region info to your Opal tap-ons data (assumed dataframe: daily_totals_by_region)
daily_totals_with_region <- daily_totals_by_region %>%
  left_join(sydney_sa2_regions %>% select(SA2_NAME, Sydney_Region), by = c("ti_region" = "SA2_NAME"))

# Aggregate Opal tap-ons by Sydney_Region for Trains only
daily_totals_by_sydney_region <- daily_totals_with_region %>%
  filter(mode_name == "Train") %>%
  group_by(Sydney_Region) %>%
  summarise(Total_Tap_Ons = sum(Total_Tap_Ons, na.rm = TRUE),
            Total_Tap_Offs = sum(Total_Tap_Offs, na.rm = TRUE),
            .groups = "drop")

# Join aggregated tap-ons to Sydney region polygons
sydney_map_data <- sydney_region_polygons %>%
  left_join(daily_totals_by_sydney_region, by = "Sydney_Region")

# Plot gradient map of Opal Train Tap-Ons by Sydney Region
ggplot(sydney_map_data) +
  geom_sf(aes(fill = Total_Tap_Ons), color = "black", size = 0.3) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90") +
  labs(
    fill = "Total Tap-Ons",
    title = "Opal Train Tap-Ons by Sydney Region",
    caption = "Data source: Opal Patronage & ABS"
  ) +
  theme_minimal()



```

```{r}
# Load libraries
library(sf)
library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)

# Step 1: Read SA2 map shapefile
sa2 <- st_read("data/sydney_map/SA2_2021_AUST_GDA2020.shp")

# Step 2: Read train entry/exit data from your single CSV file
entry_exit_data <- read_csv("data/train_data_entry_exit/aug2024_may2025.csv")

# Step 3: Clean up station names and filter for train entries
stations_to_ignore <- c("Rosehill", "Camellia", "Rydalmere", "Dundas", "Telopea", "Carlingford")

entry_exit_data <- entry_exit_data %>%
  filter(Station_Type %in% c("train", "Metro Shared")) %>%
  mutate(Train_Station = gsub(" Station", "", Station),
         Train_Station = trimws(Train_Station)) %>%
  filter(!Train_Station %in% stations_to_ignore,
         Entry_Exit == "Entry") %>%
  mutate(Trip = ifelse(Trip == "Less than 50", "50", Trip),
         Trip = as.numeric(Trip))

# Step 4: Summarize total entries per station
station_summary <- entry_exit_data %>%
  group_by(Train_Station) %>%
  summarise(Total_Entries = sum(Trip, na.rm = TRUE))

# Step 5: Load station coordinates
station_locations <- read_csv("data/TrainStationEntranceLocations/stationentrances2020_v4.csv")

# Filter to only stations in data and join coordinates
stations <- station_locations %>%
  filter(!duplicated(Train_Station)) %>%
  inner_join(station_summary, by = "Train_Station") %>%
  drop_na(LONG, LAT)

# Convert to sf object
stations_sf <- st_as_sf(stations, coords = c("LONG", "LAT"), crs = 4326)

# New bbox for tighter zoom (e.g. Sydney CBD)
bbox <- st_bbox(c(xmin = 151.1, ymin = -33.92, xmax = 151.25, ymax = -33.81), crs = st_crs(sa2))
sa2_cropped <- st_crop(sa2, bbox)

# Plot with better colouring
ggplot() +
  geom_sf(data = sa2_cropped, fill = "grey80", color = "grey40") +
  geom_sf(data = stations_sf, aes(size = Total_Entries, color = Total_Entries), alpha = 0.8) +
  scale_color_viridis_c(option = "plasma", name = "Total Entries") +
  scale_size_continuous(range = c(1, 8)) +
  coord_sf(xlim = c(151.1, 151.25), ylim = c(-33.92, -33.81)) +
  labs(
    title = "Train Station Entries Overlaid on SA2 Map",
    subtitle = "Zoomed: Sydney CBD (Aug 2024 – May 2025)",
    caption = "Data: NSW Train Stations + ABS SA2"
  ) +
  theme_minimal() +
  theme(legend.position = "right")


```

```{r}
library(sf)
library(dplyr)
library(readr)
library(ggplot2)

# --- Step 1: Load SA2 shapefile ---
sa2 <- st_read("data/sydney_map/SA2_2021_AUST_GDA2020.shp")

# --- Step 2: Load population density data ---
pop_density_data <- read_csv("data/sa2_pop.csv")  # Replace with your CSV path

# --- Step 3: Join population density to SA2 ---
sa2_pop <- sa2 %>%
  left_join(pop_density_data, by = c("SA2_NAME21" = "sa2_name"))

# --- Step 4: Load train entry/exit data ---
entry_exit_data <- read_csv("data/train_data_entry_exit/aug2024_may2025.csv")

# Stations to ignore (optional)
stations_to_ignore <- c("Rosehill", "Camellia", "Rydalmere", "Dundas", "Telopea", "Carlingford")

# Clean and filter train data
entry_exit_data <- entry_exit_data %>%
  filter(Station_Type %in% c("train", "Metro Shared")) %>%
  mutate(Train_Station = gsub(" Station", "", Station),
         Train_Station = trimws(Train_Station)) %>%
  filter(!Train_Station %in% stations_to_ignore,
         Entry_Exit == "Entry") %>%
  mutate(Trip = ifelse(Trip == "Less than 50", "50", Trip),
         Trip = as.numeric(Trip))

# Summarize total entries per station
station_summary <- entry_exit_data %>%
  group_by(Train_Station) %>%
  summarise(Total_Entries = sum(Trip, na.rm = TRUE))

# Load station locations
station_locations <- read_csv("data/TrainStationEntranceLocations/stationentrances2020_v4.csv")

# Filter & join coordinates for stations
stations <- station_locations %>%
  filter(!duplicated(Train_Station)) %>%
  inner_join(station_summary, by = "Train_Station") %>%
  filter(!is.na(LONG) & !is.na(LAT))

# Convert station data to sf object
stations_sf <- st_as_sf(stations, coords = c("LONG", "LAT"), crs = 4326)

# New bbox for tighter zoom (e.g. Sydney CBD)
bbox <- st_bbox(c(xmin = 151.1, ymin = -33.92, xmax = 151.25, ymax = -33.81), crs = st_crs(sa2))
sa2_cropped <- st_crop(sa2, bbox)

# Crop SA2 polygons to bounding box
sa2_cropped <- st_crop(sa2_pop, bbox)

# --- Step 6: Plot combined map ---
ggplot() +
  geom_sf(data = sa2_cropped, aes(fill = pop_density), color = "white", size = 0.1) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90", name = "Population Density") +
  geom_sf(data = stations_sf, aes(size = Total_Entries, color = Total_Entries), alpha = 0.8) +
  scale_color_viridis_c(option = "magma", name = "Train Entries") +
  scale_size_continuous(range = c(1, 8)) +
  coord_sf(xlim = c(bbox["xmin"], bbox["xmax"]), ylim = c(bbox["ymin"], bbox["ymax"])) +
  labs(
    title = "Population Density and Train Station Entries in Sydney SA2 Districts",
    subtitle = "Population density by district with train entries overlay",
    caption = "Data: Your CSV, NSW Train Stations & ABS SA2"
  ) +
  theme_minimal() +
  theme(legend.position = "right")

```

```{r}
# Load required libraries
library(sf)
library(dplyr)
library(readr)
library(ggplot2)

# --- Step 1: Load SA2 shapefile ---
sa2 <- st_read("data/sydney_map/SA2_2021_AUST_GDA2020.shp")

# --- Step 2: Load population density data ---
pop_density_data <- read_csv("data/sa2_pop.csv")  # Replace with your actual path

# --- Step 3: Join population density to SA2 ---
sa2_pop <- sa2 %>%
  left_join(pop_density_data, by = c("SA2_NAME21" = "sa2_name"))

# --- Step 4: Load train entry/exit data ---
entry_exit_data <- read_csv("data/train_data_entry_exit/aug2024_may2025.csv")

# Filter only train stations and clean names
stations_to_ignore <- c("Rosehill", "Camellia", "Rydalmere", "Dundas", "Telopea", "Carlingford")
entry_exit_data <- entry_exit_data %>%
  filter(Station_Type %in% c("train", "Metro Shared")) %>%
  mutate(Train_Station = gsub(" Station", "", Station),
         Train_Station = trimws(Train_Station)) %>%
  filter(!Train_Station %in% stations_to_ignore,
         Entry_Exit == "Entry") %>%
  mutate(Trip = ifelse(Trip == "Less than 50", "50", Trip),
         Trip = as.numeric(Trip))

# Summarize total entries per station
station_summary <- entry_exit_data %>%
  group_by(Train_Station) %>%
  summarise(Total_Entries = sum(Trip, na.rm = TRUE))

# --- Step 5: Load station locations ---
station_locations <- read_csv("data/TrainStationEntranceLocations/stationentrances2020_v4.csv")

# Merge with summary data
stations <- station_locations %>%
  filter(!duplicated(Train_Station)) %>%
  inner_join(station_summary, by = "Train_Station") %>%
  filter(!is.na(LONG) & !is.na(LAT))

# Convert to sf object
stations_sf <- st_as_sf(stations, coords = c("LONG", "LAT"), crs = 4326)

# --- Step 6: Load train line shapefile (must be LINESTRING format) ---
train_lines <- st_read("data/SydneyTrainRoutes/sydneytrains/SydneyTrains.shp")  # Replace with your actual path

# --- Step 7: Transform all spatial objects to same CRS (WGS84) ---
sa2_pop <- st_transform(sa2_pop, crs = 4326)
stations_sf <- st_transform(stations_sf, crs = 4326)
train_lines <- st_transform(train_lines, crs = 4326)

# --- Step 8: Crop all spatial data to Sydney CBD bbox ---
bbox <- st_bbox(c(xmin = 151.1, ymin = -33.92, xmax = 151.25, ymax = -33.81), crs = st_crs(4326))
sa2_cropped <- st_crop(sa2_pop, bbox)
stations_cropped <- st_crop(stations_sf, bbox)
train_lines_cropped <- st_crop(train_lines, bbox)

# --- Step 9: Plot combined map ---
ggplot() +
  geom_sf(data = sa2_cropped, aes(fill = pop_density), color = "white", size = 0.1) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90", name = "Population Density") +

  geom_sf(data = train_lines_cropped, color = "red", size = 1, alpha = 0.6) +

  geom_sf(data = stations_cropped, aes(size = Total_Entries, color = Total_Entries), alpha = 0.9) +
  scale_color_viridis_c(option = "magma", name = "Train Entries") +
  scale_size_continuous(range = c(1, 8)) +

  coord_sf(xlim = c(bbox["xmin"], bbox["xmax"]), ylim = c(bbox["ymin"], bbox["ymax"])) +
  labs(
    title = "Population Density and Train Station Entries in Sydney SA2 Districts",
    subtitle = "SA2 gradient by population density, with red train lines and entry volume at stations",
    caption = "Data: ABS SA2 2021, Opal Patronage 2024–2025, Station Entrances"
  ) +
  theme_minimal() +
  theme(legend.position = "right")


```

```{r}
library(sf)
library(dplyr)
library(readr)
library(ggplot2)

# Step 1: Load SA2 shapefile
sa2 <- st_read("data/sydney_map/SA2_2021_AUST_GDA2020.shp")

# Step 2: Load population density CSV
pop_density_data <- read_csv("data/sa2_pop.csv")

# Step 3: Join population density data to SA2
sa2_pop <- sa2 %>%
  left_join(pop_density_data, by = c("SA2_NAME21" = "sa2_name"))

# Step 4: Load train entry/exit data
entry_exit_data <- read_csv("data/train_data_entry_exit/aug2024_may2025.csv")

stations_to_ignore <- c("Rosehill", "Camellia", "Rydalmere", "Dundas", "Telopea", "Carlingford")

entry_exit_data <- entry_exit_data %>%
  filter(Station_Type %in% c("train", "Metro Shared")) %>%
  mutate(Train_Station = gsub(" Station", "", Station),
         Train_Station = trimws(Train_Station)) %>%
  filter(!Train_Station %in% stations_to_ignore,
         Entry_Exit == "Entry") %>%
  mutate(Trip = ifelse(Trip == "Less than 50", "50", Trip),
         Trip = as.numeric(Trip))

station_summary <- entry_exit_data %>%
  group_by(Train_Station) %>%
  summarise(Total_Entries = sum(Trip, na.rm = TRUE))

# Step 5: Load station locations
station_locations <- read_csv("data/TrainStationEntranceLocations/stationentrances2020_v4.csv")

stations <- station_locations %>%
  filter(!duplicated(Train_Station)) %>%
  inner_join(station_summary, by = "Train_Station") %>%
  filter(!is.na(LONG) & !is.na(LAT))

stations_sf <- st_as_sf(stations, coords = c("LONG", "LAT"), crs = 4326)

# Step 6: Load train line shapefile (Projected CRS)
train_lines <- st_read("data/SydneyTrainRoutes/sydneytrains/SydneyTrains.shp")

# Step 7: Transform all spatial data to EPSG:4326 (WGS84)
sa2_pop <- st_transform(sa2_pop, 4326)
stations_sf <- st_transform(stations_sf, 4326)
train_lines <- st_transform(train_lines, 4326)

# Step 8: Define bounding box (larger than before to capture lines)
bbox <- st_bbox(c(xmin = 150.90, ymin = -34.10, xmax = 151.30, ymax = -33.70), crs = st_crs(4326))

# Convert bbox to simple feature polygon for intersection filtering
bbox_sf <- st_as_sfc(bbox)

# Filter features that intersect bounding box to avoid clipping lines partially
sa2_cropped <- st_crop(sa2_pop, bbox)
stations_cropped <- st_crop(stations_sf, bbox)
train_lines_cropped <- train_lines %>% filter(st_intersects(., bbox_sf, sparse = FALSE))

# Step 9: Plot
ggplot() +
  geom_sf(data = sa2_cropped, aes(fill = pop_density), color = "white", size = 0.1) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90", name = "Population Density") +

  geom_sf(data = train_lines_cropped, color = "red", size = 1, alpha = 0.6) +

  geom_sf(data = stations_cropped, aes(size = Total_Entries, color = Total_Entries), alpha = 0.9) +
  scale_color_viridis_c(option = "magma", name = "Train Entries") +
  scale_size_continuous(range = c(1, 8)) +

  coord_sf(xlim = c(bbox["xmin"], bbox["xmax"]), ylim = c(bbox["ymin"], bbox["ymax"])) +

  labs(
    title = "Population Density and Train Station Entries in Sydney SA2 Districts",
    subtitle = "SA2 gradient by population density, with red train lines and entry volume at stations",
    caption = "Data: ABS SA2 2021, Opal Patronage 2024–2025, Station Entrances"
  ) +
  theme_minimal() +
  theme(legend.position = "right")

```
